Tree height and hydraulic traits shape growth responses across droughts in a temperate broadleaf forest

### 2020-02-10 manuscript/climate_sensitivity_draft.Rmd
working on discussion

One fundamental question regarding forest responses to drought is what drives the observed tendency for large trees to be more affected by drought. @bennett_larger_2015 was the first to show greater growth reductions for larger trees on a global scale, and numerous subsequent studies have reinforced this finding (*e.g.*, @stovall_tree_2019; @hacket-pain_consistent_2016], although there are exceptions [@zuleta_drought-induced_2017]. However, it has yet to be resolved which of several potential underlying mechanisms drive this pattern. First, taller trees have a greater biophysical challenge of lifting water greater distances against the effects of gravity and friction [@mcdowell_relationships_2011; @mcdowell_darcys_2015; @ryan_hydraulic_2006; @couvreur_water_2018], especially during drought [@zhang_size-dependent_2009]. To meet this challenge, hydraulic traits vary in relation to tree height within indivdiuals {**REFS**}, within species {**REFS**}, and across species [@liu_hydraulic_2019]. Despite the tendency for hydrualic traits to become more conservative with height on an individual tree {**REFS**}, the need for greater hydraulic efficiency (*i.e.,* larger xylem conduits) in tall trees [@liu_hydraulic_2019]{**REFS**} may leave them more vulnerable during drought {**REFS-- Mark Olsen--mentioned by Lawren**}. Second, larger trees may have lower drought resistance because of their higher crown positions, where they are exposed to higher solar radiation, greater wind speeds, and lower humidity [@koike_leaf_2001; @kunert_revised_2017 **REFS-KAT**). Subcanopy trees tend to fare better specifically due to the benefits of a buffered environment [@pretzsch_drought_2018]. Potentially counteracting the biophysical challenges faced by large trees, their larger root systems confer a potential advantage in terms of allowing greater access to water; however, it appears that this effect is usually insufficient to offset the costs of height and/or crown exposure. A final mechanism that could mediate tree size-related responses to drought is how species, and their associated hydraulic traits, are distributed with respect to size [@meakem_role_2018; @liu_hydraulic_2019]. Within a forest community, greater drought vulnerability of larger trees could occur if the larger size classes were dominated by species less adapted to handle drought. Alternatively, larger size classes may be dominated by species that are inherently better adapted to greater biophysical challenges--as is the case in tropical moist forests of Panama, where larger size classes contain greater proportions of deciduous species [@condit_quantifying_2000; @meakem_role_2018]. Understanding the mechanisms driving the greater growth reductions of larger trees during drought will require sorting out the interactive effects of height, canopy position, soil water availability, and species' traits.  {**ask Lawren to check revisions to this paragraph**}

@@

Tree size, microenvironment, and hydraulic traits shaped tree growth responses across three droughts in a temperate deciduous forest (Table 1). The greater susceptibility of larger trees to drought, similar to forests worldwide [@bennett_larger_2015], was driven primarily by their height [@liu_effect_1993; @stovall_tree_2019]. There was a marginal additional effect of crown exposure, with lowest $Rt$ among the most exposed (dominant) and suppressed trees. There was no evidence that soil water availability increased drought resistance; in contrast, trees in wetter topographic positions experienced more growth reduction (consistent with [@zuleta_drought-induced_2017; @stovall_tree_2019]), and the larger potential rooting volume of large trees provided no advantage in the drier microenvironments. Drought sensitivity was not consistently linked to species' $LMA$, $WD$, or xylem architecture, but was negatively correlated with the leaf hydraulic traits ($PLA_{dry}$, $\pi_{tlp}$) in the top overall model and the top models for two of the three individual droughts [@scoffoni_leaf_2014;@bartlett_correlations_2016; @medeiros_extensive_2019]. This is a novel finding in that $PLA_{dry}$ and $\pi_{tlp}$ have not previously been linked to drought growth responses. The direction of responses was mostly consistent across droughts, supporting the conclusion that they were driven by fundamental physiological mechanisms. However, the strengths of each predictor varied across droughts (Tables 4-5), indicating that specific drought characteristics interact with tree size, microenvironment, and traits to shape which individuals are most affected. These findings significantly advance our knowledge of the factors that confer vulnerability or resistance on trees during drought.

The droughts considered here were of a magnitude that has occurred with an average frequency of approximately one per 10-15 years (Fig. 1a, @helcoski_growing_2019) and had modest impacts on tree growth (Fig. 1b). While the majority of trees experienced reduced growth, a substantial portion had increased growth (Fig. 1b), potentially due to decreased leaf area of competitors during the drought. It is likely because of the moderate severity of these droughts, along with other factors influencing tree growth, that our best models characterize only a modest amount of variation: 11-13% for all droughts combined, and 21-26% for each individual drought (Table 5). While each of these droughts has been linked to tree mortality in the eastern United States [@druckenbrod_redefining_2019], they were substantially weaker than many of the extreme, multiannual droughts that have triggered massive tree die-off in other regions (e.g., [@allen_global_2010]). Such extremes have not occurred in the Eastern United States within the past several decades [@clark_impacts_2016; @druckenbrod_redefining_2019; @berdanier_multiyear_2016], and thus our findings--while generally consistent with findings from more severe droughts [@bennett_larger_2015; @zuleta_drought-induced_2017; @stovall_tree_2019]--apply primarily to moderate-severity droughts.

**This paragraph needs work-- it is repetitive with intro**
Our analysis indicates that height--as opposed to canopy position or soil water availability--is the primary factor through which tree size mediates drought response, and that its effect holds when also accounting for species' traits.  Taller trees face inherent biophysical challenges in lifting water a greater distance against the effects of gravity and friction [@ryan_hydraulic_2006; @mcdowell_darcys_2015; @mcdowell_relationships_2011; @couvreur_water_2018]. Vertical gradients in stem and leaf traits-including smaller and thicker leaves (higher LMA), more negative $P50$, and lower hydraulic conductivity at greater heights [@couvreur_water_2018; @koike_leaf_2001; @mcdowell_relationships_2011]-make it biophysically possible for trees to become tall [@couvreur_water_2018], yet height becomes a liability when drought incurs additional hydraulic challenges [@olson_plant_2018]. *Trade-offs can also occur, whereby traits conducive to efficient water transport may also lead to poor ability to recover from or re-route water around embolisms, for example in ponderosa pine [@roskilly_conflicting_2019].* Taller trees also face different microenvironments (Fig. 2a-b), in part because they are more likely to be in dominant canopy positions (Fig. 2d). Even under non-drought conditions, evaporative demand increases with tree height, and taller trees are more closely coupled to the atmosphere (**REFS- Jarvis 1984?**; [@bretfeld_plant_2018; @kunert_revised_2017]). Exposed canopy leaves reach higher temperatures [@smith_temperature_1977], particularly during drought when solar radiation tends to be higher and less water is available for evaporative cooling of the leaves. Correlation between height and canopy position (Fig. 2d) makes it challenging to disentangle the relative importance of height *per se* from microenvironment. However, significant decoupling between height and canopy position can be introduced by the configuration of neighboring trees (Fig. 2d) [@muller-landau_testing_2006], and we show that height is a far stronger predictor of drought response than crown position for the three droughts evaluated in our study area (Tables 1,4,5).

*Height and dominant canopy position had stronger negative effects in the 1966 drought than in the others (Tables 4-5), which could potentially be linked to the key reality that this was the longest and most severe multiannual drought in recent history, increasing the propensity for tree size to be a stronger predictor of drought response [@bennett_larger_2015; @stovall_tree_2019].* 
*In this section, be sure to acknowledge collinearity between CP and H*

@@

>>>>

One fundamental question regarding forest responses to drought is what drives the observed tendency for large trees to be more affected by drought. Greater growth reductions for larger trees was first shown on a global scale by @bennett_larger_2015, and numerous subsequent studies have reinforced this finding (*e.g.*, @stovall_tree_2019; @hacket-pain_consistent_2016), although there are exceptions [@zuleta_drought-induced_2017]. It has yet to be resolved which of several potential underlying mechanisms drive this pattern. First, taller trees have a greater biophysical challenge of lifting water greater distances against the effects of gravity and friction [@mcdowell_relationships_2011; @mcdowell_darcys_2015; @ryan_hydraulic_2006; @couvreur_water_2018], especially during drought [@zhang_size-dependent_2009]. To meet this challenge, hydraulic traits vary in relation to tree height within indivdiuals {**REFS**}, within species {**REFS**}, and across species [@liu_hydraulic_2019]. Despite the tendency for hydrualic traits to become more conservative with height on an individual tree {**REFS**}, the need for greater hydraulic efficiency (*i.e.,* larger xylem conduits) in tall trees [@liu_hydraulic_2019]{**REFS**} may leave them more vulnerable during drought {**REFS-- Mark Olsen--mentioned by Lawren**}. Second, larger trees may have lower drought resistance because of their higher crown positions, where they are exposed to higher solar radiation, greater wind speeds, and lower humidity [@koike_leaf_2001; @kunert_revised_2017 **REFS-KAT**). Subcanopy trees tend to fare better specifically due to the benefits of a buffered environment [@pretzsch_drought_2018]. Potentially counteracting the biophysical challenges faced by large trees, their larger root systems confer a potential advantage in terms of allowing greater access to water; however, it appears that this effect is usually insufficient to offset the costs of height and/or crown exposure. A final mechanism that could mediate tree size-related responses to drought is how species, and their associated hydraulic traits, are distributed with respect to size [@meakem_role_2018; @liu_hydraulic_2019]. Within a forest community, greater drought vulnerability of larger trees could occur if the larger size classes were dominated by species less adapted to handle drought. Alternatively, larger size classes may be dominated by species that are inherently better adapted to greater biophysical challenges--as is the case in tropical moist forests of Panama, where larger size classes contain greater proportions of deciduous species [@condit_quantifying_2000; @meakem_role_2018]. Understanding the mechanisms driving the greater growth reductions of larger trees during drought will require sorting out the interactive effects of height, canopy position, soil water availability, and species' traits.  {**ask Lawren to check revisions to this paragraph**}

@@

Tree size, microenvironment, and hydraulic traits shaped tree growth responses across three droughts in a temperate deciduous forest (Table 1). The greater susceptibility of larger trees to drought, similar to forests worldwide [@bennett_larger_2015], was driven primarily by their height [@liu_effect_1993; @stovall_tree_2019]. There was a marginal additional effect of crown exposure, with a tendency for lowest $Rt$ among the most exposed (dominant) and suppressed trees. There was no evidence that soil water availability increased drought resistance; in contrast, trees in wetter topographic positions had lower $Rt$ (consistent with @zuleta_drought-induced_2017; @stovall_tree_2019), and the larger potential rooting volume of large trees provided no advantage in the drier microenvironments. The negative effect of height on $Rt$ held when also accounting for species' traits.  Drought sensitivity was not consistently linked to species' $LMA$, $WD$, or xylem architecture, but was negatively correlated with the leaf hydraulic traits ($PLA_{dry}$, $\pi_{tlp}$) in the top overall model and the top models for two of the three individual droughts [@scoffoni_leaf_2014;@bartlett_correlations_2016; @medeiros_extensive_2019]. This is a novel finding in that $PLA_{dry}$ and $\pi_{tlp}$ have not previously been linked to drought growth responses. The direction of responses was mostly consistent across droughts, supporting the conclusion that they were driven by fundamental physiological mechanisms. However, the strengths of each predictor varied across droughts (Tables 4-5), indicating that specific drought characteristics interact with tree size, microenvironment, and traits to shape which individuals are most affected. These findings significantly advance our knowledge of the factors that confer vulnerability or resistance on trees during drought.

The droughts considered here were of a magnitude that has occurred with an average frequency of approximately one per 10-15 years (Fig. 1a, @helcoski_growing_2019) and had modest impacts on tree growth (Fig. 1b). While the majority of trees experienced reduced growth, a substantial portion had increased growth (Fig. 1b), potentially due to decreased leaf area of competitors during the drought. It is likely because of the moderate impact of these droughts, along with other factors influencing tree growth, that our best models characterize only a modest amount of variation: 11-13% for all droughts combined, and 21-26% for each individual drought (Table 5). *Height and dominant canopy position had stronger negative effects in the 1966 drought than in the others (Tables 4-5), potentially because dry conditions in the two preceding years (Fig. S2) severely stressed a larger portion of trees (Fig. 1b) and increased the propensity for tree size to be a stronger predictor of drought response [@bennett_larger_2015; @stovall_tree_2019].* 
**While each of these droughts has been linked to tree mortality in the eastern United States [@druckenbrod_redefining_2019], they were substantially weaker than many of the extreme, multiannual droughts that have triggered massive tree die-off in other regions (e.g., [@allen_global_2010]). Such extremes have not occurred in the Eastern United States within the past several decades [@clark_impacts_2016; @druckenbrod_redefining_2019; @berdanier_multiyear_2016], and thus our findings--while generally consistent with findings from more severe droughts [@bennett_larger_2015; @zuleta_drought-induced_2017; @stovall_tree_2019]--apply primarily to moderate-severity droughts. **

Our analysis indicates that tree height has a stronger influence on drought response than does canopy position (Tables 1,4,5). This is consistent with, and reinforces, previous findings that biophysical constraints make it impossible for trees to efficiently transport water to great heights and simultaneously maintain strong resistance and resilience to drought-induced embolism [@couvreur_water_2018; @roskilly_conflicting_2019] **REFS**. However, this result must be interpeted with some caution, given that colinearity between the two variables (Fig. **2d**) makes it impossible to confidently partition causality. Taller trees are more likely to be in dominant canopy positions (Fig. **2d**) and, largely as a consequence of their position relative to others, face different microenvironments (Fig. **2a-b**). Even under non-drought conditions, evaporative demand and maximum leaf teamperatures increase with tree height [@smith_temperature_1977; @bretfeld_plant_2018; @kunert_revised_2017], and such conditions would incur extra stress during drought, when solar radiation tends to be higher and less water is available for evaporative cooling of the leaves. However, some decoupling between height and canopy position is introduced by the configuration of neighboring trees (Fig. 2d) [@muller-landau_testing_2006], and height was an overall stronger predictor of drought response than crown position (Tables 1,4,5).

@@

**orphan text**
Taller trees face inherent biophysical challenges in lifting water a greater distance against the effects of gravity and friction [@ryan_hydraulic_2006; @mcdowell_darcys_2015; @mcdowell_relationships_2011; @couvreur_water_2018]. Vertical gradients in stem and leaf traits-including smaller and thicker leaves (higher LMA), more negative $P50$, and lower hydraulic conductivity at greater heights [@couvreur_water_2018; @koike_leaf_2001; @mcdowell_relationships_2011]-make it biophysically possible for trees to become tall [@couvreur_water_2018], yet height becomes a liability when drought incurs additional hydraulic challenges [@olson_plant_2018]. Trade-offs can also occur, whereby traits conducive to efficient water transport may also lead to poor ability to recover from or re-route water around embolisms, for example in ponderosa pine [@roskilly_conflicting_2019].
